---
title: "Exploratory Data Analysis"
author: "DATA 614 - University of Calgary"
date: "March 3, 2026"
output: pdf_document
---

# 4-plot for testing assumptions:

```{r}

library(data.table)
## another example
souvenier=fread("souvenier.csv")


souvenier=ts(souvenier,frequency = 12, start =c(1987,1)) 
# use ?ts to find about the arguments
# tells R that the time series is monthly and starts at month 1 of 1987. 
souvenier
plot(souvenier) #sequence plot

logsouv=log(souvenier)  #sometimes looking at the log values helps!

plot(logsouv)
lag.plot(souvenier,labels = FALSE,do.lines = FALSE)
hist(souvenier)

qqnorm(souvenier) # Q-Q plot for testing if the distribution is normal
qqline(souvenier) # Q-Q line for comparison
```

# Autocorrelation
With autocorrelation graphs, we are looking to see if the data is random or not. If it is random, then the data will be within the bounds. 

```{r}
acf(souvenier)

# There is some autocorrelation, meaning that the data points are correlated to the lagged values. 
# What does this mean?
```

# Box plot
Gives the IQR, as well as outliers, if any. Also looking for trends or insights.
```{r}
boxplot(souvenier)

boxplot(mpg~cyl,data=mtcars, xlab="# of Cylinders", ylab="MPG")
#could also do it this way
plot(as.factor(mtcars$cyl),mtcars$mpg)

class(mtcars$cyl)
```

# Scatter plot
Can show trends, non-randomness, changing mean and variation, ...
```{r}
plot(mtcars$mpg,mtcars$hp)
abline(lm(mtcars$hp~mtcars$mpg))

```

# Q-Q plot
```{r}
qqnorm(mtcars$mpg)
qqline(mtcars$mpg)

#looking at some basic measures

mean(mtcars$mpg)
sd(mtcars$mpg)

```

# Confidence interval
Lets look at 95% confidence intervals

```{r}
nrow(mtcars)

#Assuming that mtcars$mpg is normal: from the previous q-q plot

e=qnorm(0.975)*sd(mtcars$mpg)/sqrt(nrow(mtcars))
e

left = mean(mtcars$mpg)-e
right = mean(mtcars$mpg)+e
left
right
```

# Homoskedasticity
This is about homogeneity of the variations. Is the variation fixed?

```{r}
mtcars$mpg[1:16]
mtcars$mpg[17:32]
var.test(mtcars$mpg[1:16],mtcars$mpg[17:32])
```

# Two sample t-test
```{r}
carsttest=t.test(mtcars$mpg[1:16],mtcars$mpg[17:32], var.equal=FALSE, paired=FALSE)
carsttest
```

# Finding outliers
```{r}

ad=fread("ad.csv")

boxplot(ad)

boxplot.stats(ad$TV)$out
boxplot.stats(ad$newspaper)$out

#install.packages("outliers",repos="http://cran.us.r-project.org")
library(outliers)

#try this: ?outlier
outlier(ad$newspaper)

scores(ad$newspaper, type="z", prob=0.9)

ad$newspaper[scores(ad$newspaper, type="t", prob=0.9)]

ad$newspaper[!scores(ad$newspaper, type="t", prob=0.9)]

#using IQR
IQR(mtcars$mpg)

quantile(ad$newspaper,probs=c(.05))

ad$newspaper[ad$newspaper>quantile(ad$newspaper,probs=c(.05))&
               ad$newspaper<quantile(ad$newspaper,probs=c(.95))]

```