---
title: "Machine Learning: Neural Networks 1"
author: "University of Calgary - DATA 614"
date: "March 26, 2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, echo = TRUE)
#install.packages("formatR")
library(formatR)
rm(list=ls())
```

There are different packages available for machine learning and neural networks. Here we use the "neuralnet" package. First, a regression example.

```{r}
rm(list=ls())
# Read the Data

data = read.csv("cereals.csv", header=T)

dim(data)

# Random sampling
set.seed(1)
# creating training and test set, we use 60% of the data for training and the rest for testing
index = sample(  nrow ( data ), size = floor(0.60 * nrow(data) ), replace = FALSE)

# very important to normalize the data for NN. Normalize based on
# training set. We use a min-mix scaling here.
maxs = apply(data[index,], 2 , max)
mins = apply(data[index,], 2 , min)
scaled = data.frame(scale(data, center = mins, scale = maxs - mins))

trainNN = scaled[index , ]
testNN = scaled[-index , ]

# Fit the neural network. Need the package first
# install.packages("neuralnet")

# load library
library(neuralnet)

# fit neural network
set.seed(2)
# A NN with one hidden layer that has 3 neurons
NN = neuralnet(rating ~ calories + protein + fat + sodium + fiber, trainNN, hidden = 3 , linear.output = T ,
               lifesign = "minimal")
# The linear.output=T implies that the activation function  
# is not applied in the output layer. This is TRUE 
# for regression and FALSE for classification.
# The lifesign = "minimal" argument asks R to print out the progress. Can choose between 'none', 
# 'minimal' or 'full'. This is particularly useful when creating large models with big datasets. 
```

Now we can plot our NN model.

```{r}
plot(NN)
```

We can create another model, with 2 hidden layers, each with 3 neurons.

```{r}
# A NN with two hidden layers, each with 3 neurons
NN = neuralnet(rating ~ calories + protein + fat + sodium + fiber, trainNN, hidden = c(3,3) , linear.output = T )

# plot neural network
plot(NN)
```

Now we can use the model to make predictions.

```{r}
## Prediction using neural network
predict_testNN = compute(NN, testNN[,c(1:5)])
# Need to transform the data back to scores
predict_testNN = (predict_testNN$net.result * (maxs[6] - mins[6])) + mins[6]

plot(data$rating[-index], predict_testNN, col='blue', pch=16, ylab = "predicted rating NN", xlab = "real rating")

abline(0,1)

# Calculate Root Mean Square Error (RMSE) for the test set
RMSE.NN = sqrt(sum((data$rating[-index] - predict_testNN)^2) / nrow(data[-index,]))
RMSE.NN
```

Neural networks have many parameters that need to be decided. The best way to find a good model, is to cross-validate. Below is an example of trying different model parameters using a loop. While this does not do a cross-validation, it gives an overview of the idea. This links provides an example code for doing a k-fold cross-validation: https://www.analyticsvidhya.com/blog/2017/09/creating-visualizing-neural-network-in-r/. 

```{r}
set.seed(1)
results=data.frame()
for (j in c("logistic","tanh")){
  print(j)
  for (i in c(3,4,5)) { 
  # create the model with 1 hidden layer of size i
  NN = neuralnet(rating ~ calories + protein + fat + sodium + fiber, trainNN,   hidden = i , act.fct = j, linear.output = T )
  ## Prediction using neural network and transform the results
  predict_testNN = compute(NN, testNN[,c(1:5)])
  predict_testNN = (predict_testNN$net.result * (maxs[6] - mins[6])) + mins[6]
  
  # Calculate Root Mean Square Error (RMSE) for the test set
  RMSE.NN = sqrt(sum((data$rating[-index] - predict_testNN)^2) / nrow(data[-index,]))
  
  cat(sprintf("\"%g\" \"%f\"\n", i , RMSE.NN))
}}

```

## Classification

Neural networks can also be used for classification. Here is an example for the Activity dataset.

```{r}
rm(list=ls())
# Read the Data
data = read.csv("activity2sample.csv", header=T)

dim(data)

# load libraries
#library(neuralnet)
library(nnet) # this is needed for the class.ind() function.

# Random sampling
set.seed(1)
# creating training and test set, we use 60% of the data for training and the rest for testing
index = sample(  nrow ( data ), size = floor(0.60 * nrow(data) ), replace = FALSE)

# very important to normalize the data for NN. Normalize based on
# training set. Do not scale the classification output.
maxs = apply(data[index,1:3], 2 , max)
mins = apply(data[index,1:3], 2 , min)
scaled = data.frame(scale(data[,1:3], center = mins, scale = maxs - mins),activity=class.ind(as.factor(data[,4])))

#you can inspect the scaled values to see the 
#transformation of the class:
#View(scaled)

trainNN = scaled[index , ]
testNN = scaled[-index , ]

# fit neural network
set.seed(2)
# A NN with one hidden layer that has 3 neurons
NN = neuralnet(activity.1+activity.3+activity.4+activity.7 ~ ., trainNN, hidden = 3  , linear.output = F)

# plot neural network
plot(NN)
```

We can now use the model to make predictions.
```{r}
## Prediction using neural network
predict_testNN = compute(NN, testNN[,1:3])
# this gives us the activation values for each class.
predict_testNN$net.result

#next we need to find the classes, which is going to be the max() of the activation values
predict_testNNclass=apply(predict_testNN$net.result,1,which.max)
predict_testNNclass

#This is where you can find the class names: 
NN$model.list[1]

#need to find the actual classes (according to what we have in the model)
testActivity=max.col(testNN[,4:7])
testActivity

# Now we can calculate the accuracy and confusion table
mean(predict_testNNclass==testActivity)

table(predict_testNNclass,testActivity)
```