---
title: "Time Series Analysis - Models"
author: "University of Calgary - DATA 614"
date: "March 24, 2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, echo = TRUE)
#install.packages("formatR")
library(formatR)
rm(list=ls())
```

Here, we discuss the models used for univariate and multivariate time series data. This module is based on the Forecasting, Principles and Practice book (FPP:https://otexts.com/fpp3/), and a couple of articles (https://www.analyticsvidhya.com/blog/2015/12/complete-tutorial-time-series-modeling/ and https://towardsdatascience.com/tidy-anomaly-detection-using-r-82a0c776d523).


# Moving average
Moving average is the simplest smoothing method based on the average of the previous t observations. Is also used in classical decomposition.
```{r}
library(fpp3)
#install.packages("TTR")
library(TTR)

plot(runMean(AirPassengers,3))
plot(runMean(AirPassengers,10))

#Now using the data from fpp3

food <- aus_retail |>
  filter(Industry == "Food retailing") |>
  summarise(Turnover = sum(Turnover)) 

food |> autoplot(runMean(Turnover, 3))
food |> autoplot(runMean(Turnover, 5))
food |> autoplot(runMean(Turnover, 15))

```

# Holt-Winters (and other exponential smoothing) methods
Here is how to do Holt-winters smoothing (triple exponential smoothing). Holt-winters is also sometimes considered as a decomposition model.
```{r}
fit3=HoltWinters(log(AirPassengers))
plot(fit3) #plot the actual (in black) and fitted (red) values
summary(fit3) #these are Holt Winters model's coefficient estimates

library(forecast) #used to create the forecast
#we can use the Holt-Winters model to estimate for the next 12 months
pred=forecast(fit3, h=12)

plot(pred)
```

Another example using the fpp3 package. Here, we create a few other types of exponential smoothing approaches depending on the type of components: trend can be either additive ("A"), additive damped ("Ad"), or none ("N"); seasonality can be either additive ("A") or multiplicative ("M"). A note on damped trend: The forecasts generated by exponential smoothing methods display a constant trend (increasing or decreasing) indefinitely into the future. Empirical evidence indicates that these methods tend to over-forecast, especially for longer forecast horizons. Damped trend makes the trend flat at some point.

Also note that in in the Tidy method, creating and training models is done within the \textit{model()} function.

We will also do a validation (training and test sets).

```{r}
aus_holidays <- tourism |>
  filter(Purpose == "Holiday") |>
  summarise(Trips = sum(Trips)/1e3) 

aus_holidays |>
  autoplot(Trips)

# Training set and then create models.
aus_holidays <- tourism |>
  filter(Purpose == "Holiday") |>
  summarise(Trips = sum(Trips)/1e3) |>
  slice(1:72)

fit <- aus_holidays |>
  model(
    additive = ETS(Trips ~ error("A") + trend("A") +
                     season("A")),
    multiplicative = ETS(Trips ~ error("M") + trend("A") +
                           season("M")),
    HW1 = ETS(Trips ~ error("M") + trend("Ad") +
               season("M")),
    HW2 = ETS(Trips ~ error("A") + trend("Ad") +
               season("A"))
  )

fc <- fit |> forecast(h = 2*4)

fc |>
  autoplot(aus_holidays, level = NULL) +
  labs(title="Australian domestic tourism",
       y="Overnight trips (millions)") +
  guides(colour = guide_legend(title = "Forecast"))

```
Now let's consider the performance of the model in training data and test data.
```{r}
#training set accuracy (based on fitted values, not forecast)
aus_holidays_fc <- fit |>
  forecast(aus_holidays)

accuracy(aus_holidays_fc, aus_holidays)

aus_holidays_fc |>
autoplot(aus_holidays, level = NULL)

#test set accuracy
aus_holidays_test <- tourism |>
  filter(Purpose == "Holiday") |>
  summarise(Trips = sum(Trips)/1e3) |>
  slice(73:80)

aus_holidays_fc_test <- fit |>
  forecast(aus_holidays_test)

accuracy(aus_holidays_fc_test, aus_holidays_test)

aus_holidays_fc_test |>
autoplot(aus_holidays_test, level = NULL)
```



# Stationarity

Next, we try to achieve stationarity. Will use the "aTSA" package to be able to do the Dickey-Fuller test for non-stationary.
```{r}
#install.packages("aTSA")
library(aTSA)

#Augmented Dickey-Fuller test for non-stationary
adf.test(AirPassengers)
#As expected, this is not stationary. Lets try taking the difference of the log of data.

adf.test(diff(log(AirPassengers)))
#stationary, so we can proceed with TS analysis.

detach("package:aTSA", unload=TRUE) 
# remvoe the package aTSA so that it does not interfere with other packages.
```


There is also similar tests in the \textit{fpp3} packages. First, taking the difference and/or logs, and then doing the KPSS test (because focus is on forecasting).

```{r}
#needs differencing to remove trend
google_2015 <- gafa_stock |>
  filter(Symbol =="GOOG", year(Date) ==2015) |>
  select(Date,Close)

library(fpp3)
google_2015 |> autoplot(Close)
```

Now let's take a look at the ACF plot and ACF plot for the differences and do the Ljung-Box test for autocorrelation.
```{r}
google_2015 |> ACF(Close) |>
  autoplot() 
google_2015 |> ACF(difference(Close)) |>
  autoplot()

google_2015 |>
  mutate(diff_close = difference(Close)) |>
  features(diff_close, ljung_box, lag = 10)
```
The Ljung-Box statistic has a p-value of 0.637 (for h=10). This suggests that the daily change in the Google stock price is essentially a random amount which is uncorrelated with that of previous days.

Now run the KPSS test
```{r}
# now run the tests

google_2015 |>
  features(Close,unitroot_kpss)
```
This indicates that the null hypothesis is rejected. That is, the data is not stationary. We can difference the data, and apply the test again. There is a test to see how many diffs are needed to make it stationary.

```{r}
google_2015 |>
  features(Close,unitroot_ndiffs)
```

Now run the test on differenced data
```{r}

google_2015 |>
  features(difference(Close),unitroot_kpss)
```

This time, the p-value is reported as 0.1 (and so it could be larger than that). We can conclude that the differenced data appear stationary.

Another example which needs log transformation.
```{r}
#needs log transformation to remove heteroskedasticity
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(TotalC = sum(Cost)) |>
  mutate(Cost = TotalC / 1e6) -> a10

a10 |>
  autoplot()

a10 |>
  autoplot(log(Cost))

#to remove seasonality
a10 |>
  autoplot(
    log(Cost) |> 
  difference(12)
)
```

Now run the test on the log-transformed data, then difference of that.
```{r}
a10 |>
  features(Cost,unitroot_kpss)
#not stationary

a10 |>
  features(log(Cost),unitroot_kpss)
#not stationary

#test to see how many diffs to make it stationary
a10 |>
  features(log(Cost),unitroot_ndiffs)

#diff the data
a10 |>
  features(difference(log(Cost)),unitroot_kpss)
#Still stationary


```


# ARIMA
In order to use ARIMA, we need to decide on three parameters: the order of the AR (p), the number of times to difference the data (d), and the order of MA (q). 

Usually, a d=1 is enough for a dataset. For choosing p, we look at the ACF and for choosing q, we look at PACF. Here, I do not discuss the details of how (p,d,q) are chosen, as these require much more background, and is a bit subjective. There is a discussion of this in https://www.itl.nist.gov/div898/handbook/pmc/section4/pmc446.htm. Thankfully, there are automated ways for choosing these parameters, which we will see later. There is also some subjectivity in choosing these parameters, so it is OK to try a few ones to see which one is working better.

If the data is seasonal, we also need to decide (P,D,Q) for the seasonality effects. As a start, you can use the same (p,d,q) for seasonality effects too.

We are going to work with the log of the data, as there seems to be a (weak) exponential trend in the data (especially in variations).

```{r}
acf(diff(log(AirPassengers)))
#from acf, it seems p=0 would be OK.   

pacf(diff(log(AirPassengers)))
#from pcf, it seems q=1 would be OK.

#now we can create our ARIMA model with parameters (0,1,1) (and the same ones for seasonality effect):
fit = arima(log(AirPassengers), c(0, 1, 1),seasonal = list(order = c(0, 1, 1), period = 12))

summary(fit)

#we can use our ARIMA model to predict our TS for 10 years (10*12 months):
pred=predict(fit,10*12)

pred$pred #these are the predicted logged values for 10 years into the future.

#We can now plot our actual and predicted values:
ts.plot(AirPassengers,exp(1)^pred$pred, log = "y",  lty = c(1,3))
```

Next, we will use the auto.arima() function from library "forecast", which automatically adjusts the ARIMA parameters for us.

```{r}
#install.packages("forecast")
library(forecast)

fit2=auto.arima(log(AirPassengers))
summary(fit2)

#do the predictions with this model as well
pred=predict(fit2,10*12)

ts.plot(AirPassengers,exp(1)^pred$pred,log="y",lty = c(1,3))
```

The \textit{kpp3} package provides a better ARIMA usage.

```{r}
# Using step-wise search to find parameters
fit <- aus_holidays |>
  model(
    ARIMAstepwise = ARIMA(Trips),
  )

report(fit)

# Using a much broader search to find parameters
fit <- aus_holidays |>
  model(
    ARIMAstepwise = ARIMA(Trips),
  )

report(fit)

# If want to manually enter parameters (using the same parameters as the above models)
fit <- aus_holidays |>
  model(
    ARIMAstepwise = ARIMA(Trips ~ pdq(0,0,0)+PDQ(0,1,1)),
  )

report(fit)

fc <- fit |> forecast(h = 2*4)

fc |>
  autoplot(aus_holidays, level = NULL) +
  labs(title="Australian domestic tourism",
       y="Overnight trips (millions)") +
  guides(colour = guide_legend(title = "Forecast"))
```

It can be seen that the parameters of both models are set as ARIMA(0,0,0)(0,1,1)[4].

Now to check the performance of the model:

```{r}
aus_holidays_fc_test <- fit |>
  forecast(aus_holidays_test)

accuracy(aus_holidays_fc_test, aus_holidays_test)

aus_holidays_fc_test |>
autoplot(aus_holidays_test, level = NULL)
```

# Time series regression models

Time series regression can be done using the "tslm()" function from the "forecast" package. This allows for different ways of handling linear models on time series data. The "fpp3" package has a similar function "TSLM()"

```{r}
library(forecast)
EconData=read.csv("econAnalysis.csv")
str(EconData)
EconDataTS=ts(EconData[,2:3])
fit=tslm(Measure~Predictor, data=EconDataTS)

summary(fit)

checkresiduals(fit)
#it can be seen that this regulation does not properly capture the errors

#The Breusch-Godfrey test shows that there is autocorrelation between error terms (p-value is smaller than 0.05).

#checking the errors versus the predictor
plot(as.numeric(residuals(fit)), EconData$Predictor)

#checking the fitted values versus actual values
plot(as.numeric(fit$fitted.values), EconData$Measure)
abline(0,1)

#We can add dummy variables for seasonality and trend. However, in this dataset, there is not much of a trend or seasonality.
EconDataTS=ts(EconData[,2:3],frequency = 12)
fit2=tslm(Measure~Predictor+trend+season, data=EconDataTS)

summary(fit2)

checkresiduals(fit2)

```

Even after accounting for the trend and season, this model still suffers from autocorrelation. This is perhaps due to the lack of enough data to provide statistical power. 

We can still use this model for forecasting. An ex-ante prediction based on the last 5 periods' average value for the predictor:

```{r}
#using the fit model without trend and seasonality
fcast=forecast(fit,newdata=data.frame(Predictor=rep(mean(EconData$Predictor[72:76]),15)))
autoplot(fcast)

#using the fit2 model with trend and seasonality
fcast=forecast(fit2,newdata=data.frame(Predictor=rep(mean(EconData$Predictor[72:76]),15)))
autoplot(fcast)
```

# Linear regression with AR errors (dynamic regression)
Here, we perform regression with AR errors on a sample of economic data. The dataset includes a predictor and an economic output measure.

```{r}
B=read.csv("econAnalysis.csv")

#transform the data into time series
x=ts(B$Predictor)
y=ts(B$Measure)

# plot the scatter plot
plot.ts(x,y,xy.lines=F,xy.labels=F)

#first, we run a simple linear regression
regmodel=lm(y~x)
summary(regmodel)

#look at the autocorrelation plots: there is an autocorrelation between residuals.
acf(residuals(regmodel))

#Now with the arima errors:
regmodel2=arima(y,xreg = x,order=c(1,0,0))
summary(regmodel2)

regmodel3=auto.arima(y,xreg = x)
summary(regmodel3)

#to calculate the p-values (numerator are the coefficients:
#manually
pvalues=(1-pnorm(abs(regmodel3$coef)/sqrt(diag(regmodel3$var.coef))))*2
pvalues
#or using the lmtest package
library(lmtest)
coeftest(regmodel3)

#can check the errors using the Ljung-Box test
checkresiduals(regmodel3)

# the Ljung-Box test results show that there is no autocorrelation problem with the residuals.
```

In order to forecast, we need to divide the data to training and test, using a train-test split of the first 60 observations for training and the rest for test:

```{r}
xtrain=ts(x[1:60])

plot(ts(x[61:76]))

regmodel4=auto.arima(y[1:60],xreg = xtrain)
summary(regmodel4)

#ex-post prediction:
fcast=forecast(regmodel4, xreg=ts(x[61:76]))
autoplot(fcast)
#compared to the actual values in y[61:76]
autoplot(y)

par(mfrow=c(2, 1))
plot(fcast)
plot(y)
par(mfrow=c(1, 1))
#to calculate RMSE:

RMSE=sqrt(mean((fcast$mean-y[61:76])^2))
RMSE

MAPE = 100*mean(abs((fcast$mean-y[61:76])/y[61:76]))
MAPE
```
The above RMSE and MAPE shows that if we have access the the input variable (x), then the model performs well in predicting the output variable (y). However, if we are forecasting, we may not have access to the input variable. The method used above is called ex-post forecasting.

To make a forecast to the future, we need to set the values for x. Assuming ex-ante forecasting, there are different ways to include the input variables. For example, the input variables can be set to their last value or mean, or a smoothed value for x can be used. Below, once we use the average values of x over the last 5 days, and once an ARIMA model of x to predict into the future.

```{r}
regmodel5=auto.arima(y,xreg = x)

#ex-ante: using the average values of x
fcast=forecast(regmodel5, xreg=rep(mean(x[72:76]),15))
autoplot(fcast)

#using an arima model of x
fit = auto.arima(x)
summary(fit)
#now we use our ARIMA model to predict the x values:
pred=predict(fit,15)
pred$pred
#and use these values for x to forecast the y values into the future
fcast=forecast(regmodel5, xreg=pred$pred)
autoplot(fcast)
```

It can be seen that the ex-ante forecasting is much harder than ex-post forecasting!

We can also look at a few different scenarios:

```{r}
#increase in the measure by 5% over the next 10 years
fcastH=forecast(regmodel5, xreg=seq(mean(x[72:76]),mean(x[72:76])*1.05, 
                                    by=(mean(x[72:76])*.05)/10))
autoplot(fcastH)

#decrease in the measure by 5% over the next 10 years
fcastL=forecast(regmodel5, xreg=seq(mean(x[72:76]),mean(x[72:76])*0.95, 
                                    by=-(mean(x[72:76])*.05)/10))
autoplot(fcastL)
```

## An example with multiple input (independent) variables
Now let's consider a dataset with multiple input variables. First, try the non-time-series regression:

```{r}
data=read.csv("us_change.csv")

data = ts(data, start = 1, frequency = 4)

regmodelNonTS=lm(Unemployment~.-X-Quarter, data = data)

summary(regmodelNonTS)

acf(residuals(regmodelNonTS))
checkresiduals(regmodelNonTS)
```

Now perform the dynamic regression (with ARIMA errors). Let's use auto.arima() first

```{r}
regmodel6=auto.arima(data[,7],xreg = data[,3:6])

summary(regmodel6)
coeftest(regmodel6)
acf(residuals(regmodel6))
checkresiduals(regmodel6)
```

There is still autocorrelation in error terms. Let's play with ARIMA parameters

```{r}
regmodel7=arima(data[,7],xreg = data[,3:6],
                order = c(1, 1, 1), seasonal = list(order = c(5, 1, 1)))

summary(regmodel7)
coeftest(regmodel7)
acf(residuals(regmodel7))
checkresiduals(regmodel7)
```

This model is now OK in terms of autocorrelation of residuals, so we can trust the p-values.

Can also use the packages within \textit{fpp3} to do this more thoroughly:

```{r}
us_change |>
  autoplot(Unemployment)  +
  labs(title = "US Unemployment",
       y = "Quarterly % change")

#now fit the model with ARIMA errors
fit <- us_change |>
  model(ARIMA(Unemployment ~ Income + Production + Savings + Consumption))
report(fit)

coefficients(fit)

fit |> gg_tsresiduals()
```

The model has some issues per the ACF and Ljung-Box test above. 

Let's try a broader search setting ```stepwise = F``` in ```ARIMA()```.

```{r}
fit <- us_change |>
  model(ARIMA(Unemployment ~ Income + Production + Savings + Consumption, stepwise = F))
report(fit)

coefficients(fit)

fit |> gg_tsresiduals()
```

This is better, but not great. Let's try some other parameters:

```{r}
fit <- us_change |>
  model(ARIMA(Unemployment ~ 0 + Income + Production + Savings + Consumption + pdq(1,1,1) +PDQ(5,1,1), stepwise = F))
#the 0 is so that there is no drift. Could use 1 for drift.
report(fit)

coefficients(fit)

fit |> gg_tsresiduals()
```

# Anomaly Detection

There are several packages that enable anomaly detection in R. Here, we use the "anomalize" package. We use a tidy way of coding for this. We use STL decomposition for this.

First, we need to convert our data to a "table_time" format. For this one, I read the data from a csv file.
```{r}
APcsv=read.csv("airpassengers.csv")
APcsv$Date=as.POSIXct(APcsv$Date) #transform the Date column to a date/time format

#install.packages("tibbletime")
library(tibbletime)
AP=as_tbl_time(APcsv,index=Date) #transform the data to a tbl_time format.

library(anomalize) #tidy anomaly detection
suppressMessages(library(tidyverse)) #tidyverse packages
# The reason I added suppressMessages() before the library() was that my 
# RMarkdown LaTeX compiler was having issues with the message that comes after 
# running the tidyverse library. You don't need to add suppressMessages().

# Using STL decomposition and GESD
AP |> 
  time_decompose(AirPassengers, method = "stl", frequency = "auto", trend = "auto") |>
  anomalize(remainder, method = "gesd", alpha = 0.05, max_anoms = 0.2) |>
  plot_anomaly_decomposition()

AP |>
  time_decompose(AirPassengers) |>
  anomalize(remainder) |>
  time_recompose() |>
  plot_anomalies(time_recomposed = TRUE, ncol = 3, alpha_dots = 0.5)


#To do the same without using the tidy approach:

#decomposition = time_decompose(AP,target=AirPassengers, method = "stl", frequency = "auto", trend = "auto")
#anomalies = anomalize(decomposition, target = remainder, method = "gesd", alpha = 0.05, max_anoms = 0.2)
#plot_anomaly_decomposition(anomalies)


#decomposition = time_decompose(AP, target = AirPassengers)
#anomalies = anomalize(decomposition, target = remainder)
#recompose = time_recompose(anomalies)
#plot_anomalies(recompose, time_recomposed = TRUE, ncol = 3, alpha_dots = 0.5)

```